---
title: "NYPD Shooting Incident Data Report"
author: "Andrew See"
date: '2022-08-12'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
**NYPD Shooting Incident Data (Historic)**

List of every shooting incident that occurred in NYC going back to 2006 through the end of the previous calendar year.

This is a breakdown of every shooting incident that occurred in NYC going back to 2006 through the end of the previous calendar year. This data is manually extracted every quarter and reviewed by the Office of Management Analysis and Planning before being posted on the NYPD website. Each record represents a shooting incident in NYC and includes information about the event, the location and time of occurrence. In addition, information related to suspect and victim demographics is also included. This data can be used by the public to explore the nature of shooting/criminal activity.

## Import Libraries

```{r library, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(ggplot2)
```

## Import Data


```{r load}
data <-read_csv("https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD")
summary(data)
data
colnames(data)
```

## Tidy and Transform Data

```{r}
#remove columns that don't seem useful for analysis
df <- subset(data, select= c(OCCUR_DATE, OCCUR_TIME, BORO, PRECINCT, LOCATION_DESC, STATISTICAL_MURDER_FLAG, PERP_AGE_GROUP, PERP_SEX, PERP_RACE, VIC_AGE_GROUP, VIC_SEX, VIC_RACE))
#change NA to unknown
df
df <- df%>%
  replace_na(list(LOCATION_DESC = "Unknown", PERP_AGE_GROUP = "Unknown", PERP_SEX = "Unknown", PERP_RACE = "Unknown", VIC_AGE_GROUP = "Unknown", VIC_SEX = "Unknown", VIC_SEX = "Unknown", VIC_RACE = "Unknown"))
#change date format
df <- df %>%
  mutate(OCCUR_DATE = mdy(OCCUR_DATE))
#add year data
df$Year <- substr(df$OCCUR_DATE, 1, 4)
year_stats <- as.data.frame(table(df$Year))
year_stats$Year <- as.numeric(as.character(year_stats$Var1))
```

## Add Visualizations and Analysis

We can see that shootings were generally declining since 2006 but then beginning in 2019 there was a sharp increase.

```{r}
#add yearly plot
year_plot <- ggplot(data=year_stats, aes(x=Year,y=Freq)) + 
  geom_line(size=1) + geom_point(shape=21, colour='white', fill='black', size=5) +
  scale_y_continuous(breaks=seq(100,2100,by=100)) +
  scale_x_continuous(breaks=2006:2021) +
  theme(axis.title.x=element_blank(), axis.title.y=element_blank(),
        panel.grid.minor = element_blank()) + 
  ggtitle("NYPD Shootings per Year")
year_plot 
```

19.25% of the shootings resulted in the victim dying.
```{r}
#create pie chart of statistical murder flag
murder <- table(df$STATISTICAL_MURDER_FLAG)
murder
pct <- c(round(sum(df$STATISTICAL_MURDER_FLAG)/length(df$STATISTICAL_MURDER_FLAG),4)*100, round(1 - sum(df$STATISTICAL_MURDER_FLAG)/length(df$STATISTICAL_MURDER_FLAG),4)*100)
lbls <- paste(c("Yes", "No"),pct)
lbls <-paste(lbls, "%", sep="")
pie(murder, labels = lbls, main = "Shooting resulted in a Murder", col = rainbow(2))
```

Brooklyn has the highest number of shootings.  There doesn't seem to be much difference between boroughs if the shooting was a murder.
```{r}
#plot shootings by borough
df %>%
  ggplot(aes(x = BORO, fill = STATISTICAL_MURDER_FLAG)) +
  geom_bar() +
  labs(title = "Boroughs of New York City",
       x = "Boroughs of New York City",
       y = "Number of Shootings")
```

The highest number of shootings occur during the weekends.
```{r}
day = df %>%
  group_by(wday(OCCUR_DATE, label=TRUE)) %>%
  count()
day

day %>%
  ggplot(aes(x = `wday(OCCUR_DATE, label = TRUE)`, y = n)) +
  geom_col() +
  labs(title = "Which day of the week do the most shootings occur?",
       x = "Day of the Week",
       y = "Number of Shootings") +
  geom_bar(stat="identity", fill="blue")
```

Not surprisingly the highest number of shootings occur at night.
```{r}
hour = df %>%
  group_by(hour(hms(as.character(OCCUR_TIME)))) %>%
  count()
hour

hour %>%
  ggplot(aes(x = `hour(hms(as.character(OCCUR_TIME)))`, y = n)) +
  geom_col() +
  labs(title = "Which hour of the day do the most shootings occur?",
       x = "Hour of the Day",
       y = "Number of Shootings") +
  geom_bar(stat="identity", fill="blue")
```

Logistic regression model to predict if the shooting was a murder
```{r}
df$OCCUR_DAY = wday(df$OCCUR_DATE, label = TRUE)
df$OCCUR_HOUR = hour(hms(as.character(df$OCCUR_TIME)))
glm.fit <- glm(STATISTICAL_MURDER_FLAG ~ PERP_RACE + PERP_SEX + PERP_AGE_GROUP + VIC_AGE_GROUP + VIC_SEX + VIC_RACE + OCCUR_HOUR + OCCUR_DAY + LOCATION_DESC + BORO, data = df, family = poisson)
summary(glm.fit)
```

## Identify Bias

It would not be difficult to discriminate and stereo-type when analyzing data such as this.  Someone may have preconceived bias based on race, gender, or location.  This bias must not influence someone when analyzing the data.  Letting bias affect you, even if it is unconscious bias, may lead to presenting the data in certain ways to prove your point.  When analyzing a topic such as this it is best to use a data-driven approach and look at the data in every possible way and use an open mind.